---
title: "ARIMA_NBA"
output: html_document
date: "2025-06-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ARIMA Model for NBA Data Set
## Import Libraries and set working directory
```{r}
rm(list=ls())
while(dev.cur() > 1) dev.off()
# Set working directory
setwd("C:/Users/Besitzer/OneDrive/Dokumente/CBS_Copenhagen/Semester/SS2025/Predictive Analytics/Final Exam")
```

```{r}
library(dplyr)
library(tseries)
library(forecast)
library(ggplot2)
library(zoo)
library(reshape2)
library(tidyr)
library(broom)
library(lubridate)
```

## Load the Data and plot initally
```{r}
data <- read.csv("monthly_fg3_per_game_clean.csv", header = TRUE, sep = ",")

# plot the data to check
data$month <- as.Date(data$month)

# Basic line plot of FG3 per game over time
plot(data$month, data$FG3_per_game, type = "l", col = "blue", lwd = 2, xlab = "Year", ylab = "3PM per Game", main = "Monthly 3-Point Makes per Game Over Time")

```

## Filter our off season and data before 3 point introduction
```{r}
filtered_data <- subset(data, off_season == 0 & three_point_introduction == 1)

# Plot the filtered data
plot(filtered_data$month, filtered_data$FG3_per_game, type = "l",col = "blue", lwd = 2,xlab = "Year", ylab = "3PM per Game", main = "3PM per Game (In-Season, Post 3-Point Era)")

```

### Create subset for the eras
```{r}
intro_era         <- subset(filtered_data, shortened_three_point_line == 0 & reverted_three_point_line == 0)
short_era         <- subset(filtered_data, shortened_three_point_line == 1)
revert_era        <- subset(filtered_data, reverted_three_point_line == 1 & defensive_three_second_rule == 0 & curry_era == 0)
defenseRule_era  <- subset(filtered_data, defensive_three_second_rule == 1 & curry_era == 0)
curry_era         <- subset(filtered_data, curry_era == 1)
revert_curry_era  <- subset(filtered_data, reverted_three_point_line == 1 | defensive_three_second_rule == 1 | curry_era == 1)

```

### Create zoo objects for each era
```{r}
zoo_intro        <- zoo(intro_era$FG3_per_game, order.by = intro_era$month)
zoo_short        <- zoo(short_era$FG3_per_game, order.by = short_era$month)
zoo_revert       <- zoo(revert_era$FG3_per_game, order.by = revert_era$month)
zoo_defenseRule  <- zoo(defenseRule_era$FG3_per_game, order.by = defenseRule_era$month)
zoo_curry        <- zoo(curry_era$FG3_per_game, order.by = curry_era$month)
zoo_revert_curry <- zoo(revert_curry_era$FG3_per_game, order.by = revert_curry_era$month)
zoo_full <- zoo(filtered_data$FG3_per_game, order.by = filtered_data$month)
```

## Preprocessing
### Intro - Era
```{r}
# Plot raw time series
plot(zoo_intro, main = "FG3 per Game - Raw", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_intro <- BoxCox.lambda(zoo_intro)
zoo_intro_bc <- BoxCox(zoo_intro, lambda_intro)
plot(zoo_intro_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_intro, 2), ")"))

# ADF test on transformed series
adf_result_1 <- adf.test(zoo_intro_bc)
print(adf_result_1)

# If not stationary, difference the series
zoo_intro_diff <- diff(zoo_intro_bc)
plot(zoo_intro_diff, main = "Differenced Box-Cox Series")

# ADF test on differenced series
adf_result_2 <- adf.test(na.omit(zoo_intro_diff))
print(adf_result_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_intro_diff)), main = "ACF")
pacf(coredata(na.omit(zoo_intro_diff)), main = "PACF")
par(mfrow = c(1, 1))
```
  - ACF plot shows a strong spike at lag 1, followed by rapid drop-off and mostly insignificant lags afterward.
  
  - This suggests the presence of a Moving Average (MA) component, likely `q = 1`.
  
  - PACF plot reveals many significant negative spikes up to lag 10, then a tapering pattern.
  
  - This long tail in PACF is characteristic of an Autoregressive (AR) process with several terms.
  
  - The ADF test after differencing confirms the series is stationary.
  
  - Based on these diagnostics, the following ARIMA models are worth comparing:
  
    - ARIMA(3,1,1): Captures moderate AR structure with a simple MA term.
    
    - ARIMA(5,1,1): Allows for the longer AR tail observed in the PACF plot.
    

### Short - Era
```{r}
# Plot raw time series
plot(zoo_short, main = "FG3 per Game - Shortened Line Era (Raw)", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_short <- BoxCox.lambda(zoo_short)
zoo_short_bc <- BoxCox(zoo_short, lambda_short)
plot(zoo_short_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_short, 2), ")"))

# ADF test on transformed series
adf_result_short_1 <- adf.test(coredata(zoo_short_bc))
print(adf_result_short_1)

# If not stationary, difference the series
zoo_short_diff <- diff(zoo_short_bc)
plot(zoo_short_diff, main = "Differenced Box-Cox Series (Short Era)")

# ADF test on differenced series
adf_result_short_2 <- adf.test(coredata(na.omit(zoo_short_diff)))
print(adf_result_short_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_short_diff)), main = "ACF - Short Era")
pacf(coredata(na.omit(zoo_short_diff)), main = "PACF - Short Era")
par(mfrow = c(1, 1))
```
  - ACF plot shows a strong spike at lag 1 and then cuts off quickly, suggesting an MA(1) process.
  
  - PACF displays several small negative spikes, with a potential tail-off pattern, but no clear AR structure.
  
  - The short length of the time series may limit reliability of deeper AR or MA terms.
  
  - The ADF test confirms the differenced series is stationary.
  
  - Due to the small sample size and clear ACF cutoff, simple models are preferred:
  
    - ARIMA(0,1,1): A minimal model with only a Moving Average component.
    
    - ARIMA(1,1,1): Adds a basic AR term in case there’s minor autocorrelation in the PACF.


### Revert - Era
```{r}
# Plot raw time series
plot(zoo_revert, main = "FG3 per Game - Reverted Line Era (Raw)", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_revert <- BoxCox.lambda(zoo_revert)
zoo_revert_bc <- BoxCox(zoo_revert, lambda_revert)
plot(zoo_revert_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_revert, 2), ")"))

# ADF test on transformed series
adf_result_revert_1 <- adf.test(coredata(zoo_revert_bc))
print(adf_result_revert_1)

# If not stationary, difference the series
zoo_revert_diff <- diff(zoo_revert_bc)
plot(zoo_revert_diff, main = "Differenced Box-Cox Series (Reverted Era)")

# ADF test on differenced series
adf_result_revert_2 <- adf.test(coredata(na.omit(zoo_revert_diff)))
print(adf_result_revert_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_revert_diff)), main = "ACF - Reverted Era")
pacf(coredata(na.omit(zoo_revert_diff)), main = "PACF - Reverted Era")
par(mfrow = c(1, 1))

```
  - ACF plot shows a strong spike at lag 1 and multiple smaller positive autocorrelations up to lag 12, indicating persistence in the series.
  
  - PACF shows a clear spike at lag 1, followed by mostly insignificant or weak negative values.
  
  - This pattern suggests an ARIMA process with both autoregressive and moving average components.
  
  - The ADF test after differencing confirms the series is stationary.
  
  - Based on the diagnostic plots, two reasonable ARIMA model candidates are:
  
    - ARIMA(2,1,1): Captures short AR and MA behavior suggested by initial PACF/ACF cutoffs.
    
    - ARIMA(3,1,1): Slightly more flexible in case of lingering autocorrelation in residuals.


### Defense Rule - Era
```{r}
# Plot raw time series
plot(zoo_defenseRule, main = "FG3 per Game - Defensive 3-Second Rule Era (Raw)", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_defenseRule <- BoxCox.lambda(zoo_defenseRule)
zoo_defenseRule_bc <- BoxCox(zoo_defenseRule, lambda_defenseRule)
plot(zoo_defenseRule_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_defenseRule, 2), ")"))

# ADF test on transformed series
adf_result_defense_1 <- adf.test(coredata(zoo_defenseRule_bc))
print(adf_result_defense_1)

# If not stationary, difference the series
zoo_defenseRule_diff <- diff(zoo_defenseRule_bc)
plot(zoo_defenseRule_diff, main = "Differenced Box-Cox Series (Defense Rule Era)")

# ADF test on differenced series
adf_result_defense_2 <- adf.test(coredata(na.omit(zoo_defenseRule_diff)))
print(adf_result_defense_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_defenseRule_diff)), main = "ACF - Defense Rule Era")
pacf(coredata(na.omit(zoo_defenseRule_diff)), main = "PACF - Defense Rule Era")
par(mfrow = c(1, 1))
```
  - ACF plot shows a significant spike at lag 1 and minor positive autocorrelations persisting through multiple lags.
  
  - PACF plot exhibits a strong initial drop followed by a long tail of alternating small spikes, mostly negative.
  
  - This pattern suggests a dominant AR component with a long memory, while ACF indicates potential for mild MA
  behavior.
  - The ADF test on the differenced series confirms stationarity.
  
  - Based on these observations, two suitable ARIMA model candidates are:
  
    - ARIMA(4,1,1): Captures the decaying AR tail and a short MA component.
    
    - ARIMA(5,1,0): Focuses on the long autoregressive structure without overfitting moving average terms.


### Curry - Era
```{r}
# Plot raw time series
plot(zoo_curry, main = "FG3 per Game - Curry Era (Raw)", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_curry <- BoxCox.lambda(zoo_curry)
zoo_curry_bc <- BoxCox(zoo_curry, lambda_curry)
plot(zoo_curry_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_curry, 2), ")"))

# ADF test on transformed series
adf_result_curry_1 <- adf.test(coredata(zoo_curry_bc))
print(adf_result_curry_1)

# If not stationary, difference the series
zoo_curry_diff <- diff(zoo_curry_bc)
plot(zoo_curry_diff, main = "Differenced Box-Cox Series (Curry Era)")

# ADF test on differenced series
adf_result_curry_2 <- adf.test(coredata(na.omit(zoo_curry_diff)))
print(adf_result_curry_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_curry_diff)), main = "ACF - Curry Era")
pacf(coredata(na.omit(zoo_curry_diff)), main = "PACF - Curry Era")
par(mfrow = c(1, 1))

```
  - ACF plot shows a strong spike at lag 1 followed by a sharp drop and low values beyond, suggesting a short-term MA
  component.
  
  - PACF reveals significant negative spikes at early lags (1 to ~6), followed by a gradual decline.
  
  - This pattern points to a longer AR structure with a prominent initial response.
  
  - The ADF test confirms that the differenced series is stationary.
  
  - Based on these observations, two appropriate ARIMA model candidates are:
  
    - ARIMA(3,1,1): Balances a modest AR tail with a short MA component.
    
    - ARIMA(4,1,1): Captures the extended PACF tail more fully while maintaining a single MA term.


### Revert Curry - Era (Modern Era)
```{r}
# Plot raw time series
plot(zoo_revert_curry, main = "FG3 per Game - Modern (Raw)", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_revert_curry <- BoxCox.lambda(zoo_revert_curry)
zoo_revert_curry_bc <- BoxCox(zoo_revert_curry, lambda_revert_curry)
plot(zoo_revert_curry_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_revert_curry, 2), ")"))

# ADF test on transformed series
adf_result_rc_1 <- adf.test(coredata(zoo_revert_curry_bc))
print(adf_result_rc_1)

# If not stationary, difference the series
zoo_revert_curry_diff <- diff(zoo_revert_curry_bc)
plot(zoo_revert_curry_diff, main = "Differenced Box-Cox Series (Modern Era)")

# ADF test on differenced series
adf_result_rc_2 <- adf.test(coredata(na.omit(zoo_revert_curry_diff)))
print(adf_result_rc_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_revert_curry_diff)), main = "ACF - Modern Era")
pacf(coredata(na.omit(zoo_revert_curry_diff)), main = "PACF - Modern Era")
par(mfrow = c(1, 1))

```
  - ACF plot shows a very strong spike at lag 1 followed by rapid drop-off and minor fluctuations, indicating a strong
  Moving Average (MA) component.
  
  - PACF plot displays a long sequence of significant negative values (lags 1–10+) with gradual decay, suggesting a
  high-order Autoregressive (AR) process.
  
  - The combination of a sharp ACF cut-off and a long PACF tail is characteristic of an ARIMA model with a dominant AR
  component.
  
  - The ADF test on the differenced series confirms it is stationary.
  
  - Based on these observations, the following ARIMA models are appropriate to compare:
    
    - ARIMA(4,1,1): Captures the moderate AR tail and initial MA structure.
    
    - ARIMA(6,1,1): Allows for deeper AR structure suggested by extended PACF spikes.
    
### Full Data Set Model
```{r}
# Plot raw time series
plot(zoo_full, main = "FG3 per Game - Full (Raw)", ylab = "3PM per Game", xlab = "Date")

# Box-Cox transformation (auto lambda)
lambda_full <- BoxCox.lambda(zoo_full)
zoo_full_bc <- BoxCox(zoo_full, lambda_full)
plot(zoo_full_bc, main = paste("Box-Cox Transformed (λ =", round(lambda_full, 2), ")"))

# ADF test on transformed series
adf_result_full_1 <- adf.test(coredata(zoo_full_bc))
print(adf_result_full_1)

# If not stationary, difference the series
zoo_full_diff <- diff(zoo_full_bc)
plot(zoo_full_diff, main = "Differenced Box-Cox Series (Full)")

# ADF test on differenced series
adf_result_full_2 <- adf.test(coredata(na.omit(zoo_full_diff)))
print(adf_result_full_2)

# ACF and PACF plots
par(mfrow = c(1, 2))
acf(coredata(na.omit(zoo_full_diff)), main = "ACF - Full")
pacf(coredata(na.omit(zoo_full_diff)), main = "PACF - Full")
par(mfrow = c(1, 1))

```

    
## Model Selection and Comparison
### Intro - Era
```{r}
fit_intro_1      <- Arima(coredata(zoo_intro_bc), order = c(1,1,1))
fit_intro_2      <- Arima(coredata(zoo_intro_bc), order = c(2,1,1))
fit_intro_auto   <- auto.arima(coredata(zoo_intro_bc), lambda = lambda_intro, seasonal = FALSE)

models_intro <- list(
  arima_1_1_1 = fit_intro_1,
  arima_2_1_1 = fit_intro_2,
  auto_arima  = fit_intro_auto
)
results_intro <- bind_rows(lapply(models_intro, glance), .id = "model")
print(results_intro)
```

### Short - Era
```{r}
fit_short_1    <- Arima(coredata(zoo_short_bc), order = c(0,1,1))
fit_short_2    <- Arima(coredata(zoo_short_bc), order = c(1,1,1))
fit_short_auto <- auto.arima(coredata(zoo_short_bc), lambda = lambda_short, seasonal = FALSE)

models_short <- list(
  arima_0_1_1 = fit_short_1,
  arima_1_1_1 = fit_short_2,
  auto_arima  = fit_short_auto
)
results_short <- bind_rows(lapply(models_short, glance), .id = "model")
print(results_short)
```

### Revert - Era
```{r}
fit_revert_1    <- Arima(coredata(zoo_revert_bc), order = c(2,1,1))
fit_revert_2    <- Arima(coredata(zoo_revert_bc), order = c(3,1,1))
fit_revert_auto <- auto.arima(coredata(zoo_revert_bc), lambda = lambda_revert, seasonal = FALSE)

models_revert <- list(
  arima_2_1_1 = fit_revert_1,
  arima_3_1_1 = fit_revert_2,
  auto_arima  = fit_revert_auto
)
results_revert <- bind_rows(lapply(models_revert, glance), .id = "model")
print(results_revert)

```

### Defense Rule - Era
```{r}
fit_defense_1    <- Arima(coredata(zoo_defenseRule_bc), order = c(4,1,1))
fit_defense_2    <- Arima(coredata(zoo_defenseRule_bc), order = c(5,1,0))
fit_defense_auto <- auto.arima(coredata(zoo_defenseRule_bc), lambda = lambda_defenseRule, seasonal = FALSE)

models_defense <- list(
  arima_4_1_1 = fit_defense_1,
  arima_5_1_0 = fit_defense_2,
  auto_arima  = fit_defense_auto
)
results_defense <- bind_rows(lapply(models_defense, glance), .id = "model")
print(results_defense)

```

### Curry - Era
```{r}
fit_curry_1    <- Arima(coredata(zoo_curry_bc), order = c(3,1,1))
fit_curry_2    <- Arima(coredata(zoo_curry_bc), order = c(4,1,1))
fit_curry_auto <- auto.arima(coredata(zoo_curry_bc), lambda = lambda_curry, seasonal = FALSE)

models_curry <- list(
  arima_3_1_1 = fit_curry_1,
  arima_4_1_1 = fit_curry_2,
  auto_arima  = fit_curry_auto
)
results_curry <- bind_rows(lapply(models_curry, glance), .id = "model")
print(results_curry)

```

### Modern - Era
```{r}
fit_modern_1    <- Arima(coredata(zoo_revert_curry_bc), order = c(6,1,1))
fit_modern_2    <- Arima(coredata(zoo_revert_curry_bc), order = c(8,1,1))
fit_modern_auto <- auto.arima(coredata(zoo_revert_curry_bc), lambda = lambda_revert_curry, seasonal = FALSE)

models_modern <- list(
  arima_6_1_1 = fit_modern_1,
  arima_8_1_1 = fit_modern_2,
  auto_arima  = fit_modern_auto
)
results_modern <- bind_rows(lapply(models_modern, glance), .id = "model")
print(results_modern)

```
### Full Data Set Model
```{r}
fit_full_1    <- Arima(coredata(zoo_full_bc), order = c(6,1,1))
fit_full_2    <- Arima(coredata(zoo_full_bc), order = c(8,1,1))
fit_full_auto <- auto.arima(coredata(zoo_full_bc), lambda = lambda_full, seasonal = FALSE)

models_full <- list(
  arima_6_1_1 = fit_full_1,
  arima_8_1_1 = fit_full_2,
  auto_arima  = fit_full_auto
)
results_full <- bind_rows(lapply(models_full, glance), .id = "model")
print(results_full)

```



### Summary of Results
- **Intro Era**:  
  The ARIMA(1,1,1) model performs best with the lowest AIC (32.22) and BIC (40.68), and a low sigma (0.1595). ARIMA(2,1,1) has slightly higher AIC/BIC, while `auto.arima()` performs poorly with a much higher AIC (451.56) and sigma (1.455).  
  **Selected model:** `ARIMA(1,1,1)`

- **Shortened Line Era**:  
  ARIMA(0,1,1) has the lowest AIC (-122.28) and BIC (-120.01), and a lower sigma than `auto.arima`. Although ARIMA(1,1,1) has a lower sigma, its AIC/BIC are worse. `auto.arima()` again performs worse.  
  **Selected model:** `ARIMA(0,1,1)`

- **Reverted Era**:  
  The `auto.arima()` model clearly outperforms manual models, with the best AIC (-55.61), BIC (-52.75), and lowest sigma (0.0940). Manual models like ARIMA(2,1,1) and (3,1,1) are notably worse.  
  **Selected model:** `auto.arima()`

- **Defensive Rule Era**:  
  Again, `auto.arima()` is the best performer with the lowest AIC (-213.55), BIC (-208.04), and the smallest sigma (0.0951). Manual models such as ARIMA(4,1,1) and (5,1,0) have higher AIC/BIC and sigma.  
  **Selected model:** `auto.arima()`

- **Curry Era**:  
  ARIMA(3,1,1) is preferred with the lowest AIC (54.19) and sigma (0.0072). ARIMA(4,1,1) performs slightly worse in both AIC and BIC. The `auto.arima()` model performs poorly again with an extremely high AIC (130.36) and sigma (0.7545).  
  **Selected model:** `ARIMA(3,1,1)`

- **Modern Era (Revert + Defense + Curry)**:  
  ARIMA(6,1,1) is the top choice with the best AIC (80.62), BIC (107.09), and lowest sigma (0.0814). ARIMA(4,1,1) is a weaker alternative. The `auto.arima()` model again shows poor performance with very high AIC (394.40) and sigma (0.6345).  
  **Selected model:** `ARIMA(6,1,1)`

### Fit each model on its respective era
```{r}
# 1) Intro Era --------------------------------------------------------------

# Fit on the numeric Box–Cox series
fit_intro <- Arima(coredata(zoo_intro_bc), order = c(1,1,1))

# Extract & back‐transform fitted values
back_fitted_i <- InvBoxCox(fitted(fit_intro), lambda_intro)

# Align into zoo with the exact original dates
fitted_intro_zoo <- zoo(back_fitted_i, index(zoo_intro_bc))

# Plot
plot(zoo_intro, col="black", lwd=1.5,
     main="Intro Era: Actual vs Fitted", ylab="3PM per Game", xlab="Date")
lines(fitted_intro_zoo, col="red", lwd=1.5)
legend("topleft", legend=c("Actual","Fitted"), col=c("black","red"), lty=1)



# 2) Shortened Line Era -----------------------------------------------------

fit_short <- Arima(coredata(zoo_short_bc), order = c(0,1,1))
back_fitted_s <- InvBoxCox(fitted(fit_short), lambda_short)
fitted_short_zoo <- zoo(back_fitted_s, index(zoo_short_bc))

plot(zoo_short, col="black", lwd=1.5,
     main="Shortened Line Era: Actual vs Fitted", ylab="3PM per Game", xlab="Date")
lines(fitted_short_zoo, col="red", lwd=1.5)



# 3) Reverted Era ------------------------------------------------------------

fit_revert <- auto.arima(coredata(zoo_revert_bc), lambda=lambda_revert, seasonal=FALSE)
back_fitted_r <- InvBoxCox(fitted(fit_revert), lambda_revert)
fitted_revert_zoo <- zoo(back_fitted_r, index(zoo_revert_bc))

plot(zoo_revert, col="black", lwd=1.5,
     main="Reverted Era: Actual vs Fitted", ylab="3PM per Game", xlab="Date")
lines(fitted_revert_zoo, col="red", lwd=1.5)



# 4) Defensive 3-Second Rule Era -------------------------------------------

fit_defense <- auto.arima(coredata(zoo_defenseRule_bc), lambda=lambda_defenseRule, seasonal=FALSE)
back_fitted_d <- InvBoxCox(fitted(fit_defense), lambda_defenseRule)
fitted_defense_zoo <- zoo(back_fitted_d, index(zoo_defenseRule_bc))

plot(zoo_defenseRule, col="black", lwd=1.5,
     main="Defensive 3-Second Rule Era: Actual vs Fitted", ylab="3PM per Game", xlab="Date")
lines(fitted_defense_zoo, col="red", lwd=1.5)



# 5) Curry Era --------------------------------------------------------------

fit_curry <- Arima(coredata(zoo_curry_bc), order = c(3,1,1))
back_fitted_c <- InvBoxCox(fitted(fit_curry), lambda_curry)
fitted_curry_zoo <- zoo(back_fitted_c, index(zoo_curry_bc))

plot(zoo_curry, col="black", lwd=1.5,
     main="Curry Era: Actual vs Fitted", ylab="3PM per Game", xlab="Date")
lines(fitted_curry_zoo, col="red", lwd=1.5)

```
### Combine fitted values
```{r}
# Combine all actuals and fitted segment series into one full zoo and plot together

# 1) Build the full actual series
full_actual <- zoo(filtered_data$FG3_per_game, order.by = filtered_data$month)

# 2) Combine each fitted segment into one big zoo object
all_fitted <- do.call(merge, list(
  fitted_intro_zoo,
  fitted_short_zoo,
  fitted_revert_zoo,
  fitted_defense_zoo,
  fitted_curry_zoo
))
# Rename columns
colnames(all_fitted) <- c("Intro","Short","Revert","Defense","Curry")

# 3) Plot
plot(full_actual, col = "black", lwd = 1.5,
     main = "Actual vs EraSpecific Fitted Values (ARIMA)",
     ylab = "3PM per Game", xlab = "Date")

# Add each fitted line
cols <- c("red","blue","green","purple","orange")
for(i in seq_len(ncol(all_fitted))) {
  lines(all_fitted[,i], col = cols[i], lwd = 1.5)
}

# 4) Legend
legend("topleft",
       legend = c("Actual", colnames(all_fitted)),
       col    = c("black", cols),
       lty    = 1,
       lwd    = 1.5)

```
## Test Models on unseen data (Using Test-Train Split)
```{r}
# 1) Split at 2019-10-01
train_data <- subset(filtered_data, month < as.Date("2019-10-01"))
test_data  <- subset(filtered_data, month >= as.Date("2019-10-01"))
# Convert test to zoo objects
zoo_test  <- zoo(test_data$FG3_per_game, order.by = test_data$month)

# 2) Curry‐Only Model
curry_train_df <- subset(train_data, curry_era == 1)
z_curry_tr     <- zoo(curry_train_df$FG3_per_game, order.by = curry_train_df$month)
lam_curry      <- BoxCox.lambda(z_curry_tr)
ts_curry_tr    <- coredata(BoxCox(z_curry_tr, lam_curry))
fit_curry_tr   <- Arima(ts_curry_tr, order = c(3,1,1), include.drift=TRUE)
fc_curry       <- forecast(fit_curry_tr, h = nrow(test_data))
# back‐transform and wrap as zoo
zoo_fc_curry <- zoo(
  InvBoxCox(as.numeric(fc_curry$mean), lam_curry),
  order.by = test_data$month
)

# 3) Modern‐Era Model (Revert + Defense + Curry)
modern_train_df <- subset(train_data,
  reverted_three_point_line == 1 |
  defensive_three_second_rule == 1 |
  curry_era == 1
)
z_mod_tr      <- zoo(modern_train_df$FG3_per_game, order.by = modern_train_df$month)
lam_modern    <- BoxCox.lambda(z_mod_tr)
ts_mod_tr     <- coredata(BoxCox(z_mod_tr, lam_modern))
fit_modern_tr <- Arima(ts_mod_tr, order = c(8,1,1), include.drift=TRUE)
fc_modern     <- forecast(fit_modern_tr, h = nrow(test_data))
zoo_fc_modern <- zoo(
  InvBoxCox(as.numeric(fc_modern$mean), lam_modern),
  order.by = test_data$month
)

#Full Data Model

# 1) Build the zoo on the training portion
z_full_tr  <- zoo(train_data$FG3_per_game, order.by = train_data$month)

# 2) Choose Box–Cox λ for that training series
lam_full   <- BoxCox.lambda(z_full_tr)

# 3) Apply Box–Cox & pull out the numeric vector
ts_full_tr <- coredata(BoxCox(z_full_tr, lam_full))

# 4) Fit ARIMA(8,1,1)+drift on that numeric vector
fit_full_tr <- Arima(ts_full_tr, order = c(8,1,1), include.drift = TRUE)

# 5) Forecast h = length of your test set
fc_full     <- forecast(fit_full_tr, h = nrow(test_data))

# 6) Invert back to original scale and wrap into zoo over the test dates
zoo_fc_full <- zoo(
  InvBoxCox(as.numeric(fc_full$mean), lam_full),
  order.by = test_data$month
)

# 4) Prepare data frames for ggplot
# Historical full series
df_hist <- data.frame(
  date  = filtered_data$month,
  value = filtered_data$FG3_per_game,
  series = "Historical"
)

# Curry‐only forecast (test period)
df_curry_fc <- data.frame(
  date  = index(zoo_fc_curry),
  value = coredata(zoo_fc_curry),
  series = "Forecast_Curry"
)

# Modern‐era forecast (test period)
df_modern_fc <- data.frame(
  date  = index(zoo_fc_modern),
  value = coredata(zoo_fc_modern),
  series = "Forecast_Modern"
)

# Full Data Model
df_full_fc <- data.frame(
  date  = index(zoo_fc_full),
  value = coredata(zoo_fc_full),
  series = "Forecast_Full"
)


# 5) Combine and plot
df_plot <- rbind(df_hist, df_curry_fc, df_modern_fc, df_full_fc)

ggplot(df_plot, aes(x = date, y = value, color = series)) +
  geom_line(data = subset(df_plot, series == "Historical"),     size = 0.8) +
  geom_line(data = subset(df_plot, series == "Forecast_Modern"), size = 0.8) +
  geom_line(data = subset(df_plot, series == "Forecast_Curry"),  size = 0.8) +
  geom_line(data = subset(df_plot, series == "Forecast_Full"), size = 0.8) +
  scale_color_manual(
    values = c(
      Historical      = "grey",
      Forecast_Modern = "green",
      Forecast_Curry  = "blue",
      Forecast_Full   = "orange"
    ),
    labels = c(
      Historical      = "Historical/Actual",
      Forecast_Modern = "ModernEra Forecast",
      Forecast_Curry  = "CurryOnly Forecast",
      Forecast_Full   = "FullData Forecast"
    )
  ) +
  labs(
    title = "FG3 per Game: Historical vs Forecasts (ARIMA)",
    x     = "Date",
    y     = "3PM per Game",
    color = ""   # removes the legend title
  ) +
  theme_minimal()

```
### Check Paramters
```{r}
summary(fit_curry_tr)
summary(fit_modern_tr)
summary(fit_full_tr)
```



### Check residuals of the models
```{r}
checkresiduals(fit_curry_tr)
checkresiduals(fit_modern_tr)
checkresiduals(fit_full_tr)
```
  - **Curry‐Only Model (ARIMA(3,1,1) with drift)**  
  
    - Ljung–Box test: Q* = 3.45, df = 4, p = 0.486 → no evidence of residual autocorrelation. 
    
    - Residuals fluctuate randomly around zero with no visible trend or clustering.  
    
    - ACF of residuals: all lags within confidence bounds, confirming whiteness.  
    
    - Histogram: roughly bell‐shaped and centered at zero, indicating approximate normality.  
    
    - **Conclusion:** Diagnostics support that the ARIMA(3,1,1)+drift model is well‐specified for the Curry era.
  
  - **Modern‐Era Model (ARIMA(6,1,1) with drift)**  
  
    - Ljung–Box test: Q* = 12.23, df = 3, p = 0.0067 → significant autocorrelation remains.  
    
    - Residual time plot: shows occasional spikes and minor cyclic patterns.  
    
    - ACF of residuals: several spikes exceed the confidence bounds (e.g. at lags ~9 and ~16), indicating unmodeled
    structure.  
    - Histogram: heavier tails than normal, with more extreme residuals. 
    
    - **Conclusion:** The ARIMA(6,1,1)+drift model leaves autocorrelation in the residuals—consider adding higher‐order
    terms.  

## Calculate accuracy metrics
```{r}
# Extract the numeric actuals from the zoo
actual_test <- coredata(zoo_test)

# Compute accuracy
acc_curry  <- accuracy(fc_curry,  actual_test)
acc_modern <- accuracy(fc_modern, actual_test)
acc_full   <- accuracy(fc_full,   actual_test)

# Combine into one data frame with model labels
acc_comparison <- rbind(
  data.frame(Model = "Curry-Only",    acc_curry),
  data.frame(Model = "Modern-Era",    acc_modern),
  data.frame(Model = "Full Data Set", acc_full)
)

# Print the comparison
print(acc_comparison)
```
**Comparison & Interpretation**  

  - Both ARIMA models achieve near‐zero errors in‐sample but catastrophically overfit, as test RMSE/MAE jump by two   
  orders of magnitude.  
  
  - The Modern‐Era ARIMA outperforms the Curry‐Only ARIMA on every out‐of‐sample metric (lower RMSE, MAE, MAPE and
  MASE).  
  - The enormous MASE values show that even the better model is far worse than a naïve seasonal forecast.  
 
  - If forced to choose, use the Modern-Era for slightly better hold-out performance—but overall, neither model is
  reliable for forecasting.  






