---
title: "ETS_NBA"
output:
  pdf_document: default
  html_document: default
date: "2025-06-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# ETS Model for NBA Data Set
## Import Libraries and set working directory
```{r}
library(dplyr)
library(tseries)
library(forecast)
library(ggplot2)
library(zoo)
library(reshape2)
library(seastests)



rm(list=ls())
while(dev.cur() > 1) dev.off()
# Set working directory
setwd("C:/Users/Besitzer/OneDrive/Dokumente/CBS_Copenhagen/Semester/SS2025/Predictive Analytics/Final Exam")
```

## Load the Data and plot initally
```{r}
data <- read.csv("monthly_fg3_per_game_clean.csv", header = TRUE, sep = ",")

# plot the data to check
data$month <- as.Date(data$month)

# Basic line plot of FG3 per game over time
plot(data$month, data$FG3_per_game, type = "l", col = "blue", lwd = 2, xlab = "Year", ylab = "3PM per Game", main = "Monthly 3-Point Makes per Game Over Time")

```

## Filter our off season and data before 3 point introduction
```{r}
filtered_data <- subset(data, off_season == 0 & three_point_introduction == 1)

# Plot the filtered data
plot(filtered_data$month, filtered_data$FG3_per_game, type = "l",col = "blue", lwd = 2,xlab = "Year", ylab = "3PM per Game", main = "3PM per Game (In-Season, Post 3-Point Era)")

```
## Fit ETS models for the time periods (using the dummy variables)

### Create subset for the eras
```{r}
intro_era         <- subset(filtered_data, shortened_three_point_line == 0 & reverted_three_point_line == 0)
short_era         <- subset(filtered_data, shortened_three_point_line == 1)
revert_era        <- subset(filtered_data, reverted_three_point_line == 1 & defensive_three_second_rule == 0 & curry_era == 0)
defenseRule_era  <- subset(filtered_data, defensive_three_second_rule == 1 & curry_era == 0)
curry_era         <- subset(filtered_data, curry_era == 1)
revert_curry_era  <- subset(filtered_data, reverted_three_point_line == 1 | defensive_three_second_rule == 1 | curry_era == 1)
```

### Convert each subset to a time series
```{r}
ts_filtered <- ts(filtered_data$FG3_per_game, start = c(1979, 10), frequency = 8)
ts_intro   <- ts(intro_era$FG3_per_game,   start = c(1979, 10), frequency = 8)
ts_short   <- ts(short_era$FG3_per_game,   start = c(1994, 10), frequency = 8)
ts_revert  <- ts(revert_era$FG3_per_game,  start = c(1997, 10), frequency = 8)
ts_defense <- ts(defenseRule_era$FG3_per_game, start = c(1997, 10), frequency = 8)
ts_curry   <- ts(curry_era$FG3_per_game,   start = c(2015, 10),  frequency = 8)
ts_revert_curry <- ts(revert_curry_era$FG3_per_game, start = c(1997, 10), frequency = 8)

```

```{r}
decomp_diff <- stl(ts_filtered, s.window = "periodic")
plot(decomp_diff)
print(isSeasonal(ts_filtered))
```
```{r}
seasonplot(ts_filtered, col = rainbow(20), year.labels = TRUE, main = "Seasonal Pattern")
```



### Fit ETS models for each era
```{r}
fit_intro  <- ets(ts_intro)
fit_short  <- ets(ts_short)
fit_revert <- ets(ts_revert)
fit_defense <- ets(ts_defense)
fit_curry  <- ets(ts_curry)
```

### Combine fitted values into one full timeline
```{r}
# 1) Full actual series as zoo
full_actual <- zoo(
  filtered_data$FG3_per_game,
  order.by = filtered_data$month
)

# 2) Turn each ETS fit (ts) into a zoo, using the same dates you used to build the ts
fitted_intro_zoo   <- zoo(as.numeric(fitted(fit_intro)),   order.by = intro_era$month)
fitted_short_zoo   <- zoo(as.numeric(fitted(fit_short)),   order.by = short_era$month)
fitted_revert_zoo  <- zoo(as.numeric(fitted(fit_revert)),  order.by = revert_era$month)
fitted_defense_zoo <- zoo(as.numeric(fitted(fit_defense)), order.by = defenseRule_era$month)
fitted_curry_zoo   <- zoo(as.numeric(fitted(fit_curry)),   order.by = curry_era$month)

# 3) Merge into one multi‐column zoo
all_fitted <- do.call(merge, list(
  fitted_intro_zoo,
  fitted_short_zoo,
  fitted_revert_zoo,
  fitted_defense_zoo,
  fitted_curry_zoo
))
colnames(all_fitted) <- c("Intro","Short","Revert","Defense","Curry")

# 4) Plot actual + each fitted segment
plot(
  full_actual,
  col  = "black",
  lwd  = 1.5,
  main = "Actual vs EraSpecific Fitted Values (ETS)",
  xlab = "Date",
  ylab = "3PM per Game"
)
cols <- c("red","blue","green","purple","orange")
for(i in seq_len(ncol(all_fitted))) {
  lines(all_fitted[,i], col = cols[i], lwd = 1.5)
}

# 5) Legend
legend(
  "topleft",
  legend = c("Actual", colnames(all_fitted)),
  col    = c("black", cols),
  lty    = 1,
  lwd    = 1.5,
  bty    = "n"
)

```
## Test model on a unseen data set (using train test split)

We are testing two version of model here

  - One that only uses the Curry era data
  
  - One that uses the Modern Era

```{r}
## 1) Split the raw data frame
train_df <- filtered_data %>% filter(month < as.Date("2019-10-01"))
test_df  <- filtered_data %>% filter(month >= as.Date("2019-10-01"))
h        <- nrow(test_df)

## 2) Build ts for each era on the TRAIN portion only
ts_curry_tr  <- ts(train_df$FG3_per_game[train_df$curry_era == 1],
                   start = c(2015,10), frequency = 8)
ts_modern_tr <- ts(train_df$FG3_per_game[
                     train_df$reverted_three_point_line == 1 |
                     train_df$defensive_three_second_rule == 1 |
                     train_df$curry_era == 1
                   ],
                   start = c(1997,10), frequency = 8)
ts_full_tr  <- ts(train_df$FG3_per_game, start = c(1979,10), frequency = 8)

## 3) Fit ETS and forecast
fit_curry_ets  <- ets(ts_curry_tr)
fit_modern_ets <- ets(ts_modern_tr)
fit_full_ets  <- ets(ts_full_tr)

fc_curry  <- forecast(fit_curry_ets,  h = h, level = c(80, 95))
fc_modern <- forecast(fit_modern_ets, h = h, level = c(80, 95))
fc_full  <- forecast(fit_full_ets,  h = h, level = c(80, 95))

## 4) Wrap forecasts as zoo on the test_df dates
zoo_fc_curry  <- zoo(as.numeric(fc_curry$mean),  order.by = test_df$month)
zoo_fc_modern <- zoo(as.numeric(fc_modern$mean), order.by = test_df$month)
zoo_fc_full  <- zoo(as.numeric(fc_full$mean),  order.by = test_df$month)

# 4) Prepare data frames for ggplot
# Historical full series
df_hist <- data.frame(
  date  = filtered_data$month,
  value = filtered_data$FG3_per_game,
  series = "Historical"
)

# Curry‐only forecast (test period)
df_curry_fc <- data.frame(
  date  = index(zoo_fc_curry),
  value = coredata(zoo_fc_curry),
  series = "Forecast_Curry"
)

# Modern‐era forecast (test period)
df_modern_fc <- data.frame(
  date  = index(zoo_fc_modern),
  value = coredata(zoo_fc_modern),
  series = "Forecast_Modern"
)
# Full‐era forecast (test period)
df_full_fc <- data.frame(
  date  = index(zoo_fc_full),
  value = coredata(zoo_fc_full),
  series = "Forecast_Full"
)

# 5) Combine and plot
df_plot <- rbind(df_hist, df_curry_fc, df_modern_fc, df_full_fc)

ggplot(df_plot, aes(x = date, y = value, color = series)) +
  geom_line(data = subset(df_plot, series == "Historical"),     size = 0.8) +
  geom_line(data = subset(df_plot, series == "Forecast_Modern"), size = 0.8) +
  geom_line(data = subset(df_plot, series == "Forecast_Curry"),  size = 0.8) +
  geom_line(data = subset(df_plot, series == "Forecast_Full"),    size = 0.8) +
  scale_color_manual(
    values = c(
      Historical      = "grey",
      Forecast_Modern = "green",
      Forecast_Curry  = "blue",
      Forecast_Full   = "orange"
    ),
    labels = c(
      Historical      = "Historical/Actual",
      Forecast_Modern = "ModernEra Forecast",
      Forecast_Curry  = "CurryOnly Forecast",
      Forecast_Full   = "FullEra Forecast"
    )
  ) +
  labs(
    title = "FG3 per Game: Historical vs Forecasts (ETS)",
    x     = "Date",
    y     = "3PM per Game",
    color = ""   # removes the legend title
  ) +
  theme_minimal()
```

### Check ETS Parameters
```{r}
summary(fit_curry_ets)
summary(fit_modern_ets)
summary(fit_full_ets)
```




### Check residuals of the models
```{r}
checkresiduals(fit_curry_ets)
checkresiduals(fit_modern_ets)
checkresiduals(fit_full_ets)
```
Residual diagnostics: Curry‐Era (ETS(M,A,N))

  - Ljung–Box: Q* = 4.96, df = 8, p = 0.762 → no evidence of autocorrelation.
  
  - ACF: all lags within ±2·SE, indicating whiteness.
  
  - Time plot: residuals randomly fluctuate around zero, no trend or seasonality.
  
  - Histogram: approximately Gaussian, centered at zero.
  
  - Conclusion: residuals behave like white noise, model is well‐specified.

Residual diagnostics: Modern‐Era (ETS(M,A,N))

  - Ljung–Box: Q* = 30.67, df = 24, p = 0.1635 → again, no significant autocorrelation.
  
  - ACF: small long‐lag spikes but still within confidence bounds.

  - Time plot: no systematic pattern, residuals hover around zero.

  - Histogram: roughly normal, with slight heavy tails.

  - Conclusion: residuals pass the usual checks
  
## Calculate accuracy metrics
```{r}
# 1) Extract the actuals from the test set
actual_test <- test_df$FG3_per_game

# 2) Compute accuracy for each ARIMA forecast
acc_curry_ets  <- accuracy(fc_curry,  actual_test)
acc_modern_ets <- accuracy(fc_modern, actual_test)
acc_full_ets  <- accuracy(fc_full,  actual_test)

# 3) Combine into one table with model labels
acc_comparison_ets <- rbind(
  data.frame(Model = "Curry-Only", acc_curry_ets),
  data.frame(Model = "Modern-Era", acc_modern_ets),
  data.frame(Model = "Full-Era", acc_full_ets)
)

# 4) Print the comparison
print(acc_comparison_ets)
```
**Comparison & conclusions**  

  - The Modern-Era model cuts test RMSE and MAE by roughly half versus the Curry-Only model.  
  
  - Test MAPE and MASE for Modern-Era are also less than half of the Curry-Only values. 
  
  - Curry-Only generalizes poorly (test errors high), whereas Modern-Era delivers much tighter out-of-sample accuracy.
  
  - Training residual ACF1 is near zero (−0.013) for Modern-Era, indicating more white-noise residuals than Curry-Only
  (ACF1 = 0.156).  
  
**Recommendation:** use the Modern-Era model for forecasting, as it clearly outperforms the Curry-Only approach on held-out data.  
